---
title: "class12"
author: "Laurie Chang A16891192"
format: pdf
editor: visual
---

##Section 4 Population Scale Analysis

>Q13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

There are 108 individuals who have the A/A genotype, 233 individuals with the A/G genotype, and 121 individuals with the G/G genotype.

```{r}
median_expression <- aggregate(expr$exp ~ expr$geno, data = expr, FUN = median)
median_expression
```

The median expression level for the A/A genotype is 31.24847.
The median expression level for the A/G genotype is 25.06486.
The median expression level for the G/G genotype is 20.07363.

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Let's make a boxplot!

```{r}
library(ggplot2)

ggplot(expr) + 
  aes(x = geno, y = exp, fill = geno) +
  geom_boxplot(notch = TRUE)
 
```

I can infer that there is a significant decrease in expression in the G/G phenotype in comparison to the A/A phenotype. It appears the SNP does effect the expression of ORMDL3. 

