---
title: "class18"
author: "Laurie Chang A16891192"
format: pdf
editor: visual
---

First we will examine and explore Pertussis case numbers in the US as tracked by the CDC:
https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

We can use the datapasta package to scrape this data from the website into R:

```{r}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116)
       )
```

```{r}
head(cdc)
```

>Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.
Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

I want a plot of cases per year with ggplot

```{r}
library(ggplot2)
```

```{r}
ggplot(cdc) +
  aes(cdc$year, cdc$cases) +
  geom_line() +
  geom_vline(xintercept = 1947, col = "blue") +
  geom_vline(xintercept = 1992, col = "red") +
  geom_vline(xintercept = 2020, col = "orange")

```

I notice that there is a significant decrease in the number of cases after the 1946 introduction of the wP vaccine. After the switch in 1996 to the aP vaccine, I notice that there is a spike in the number of cases. 

>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

There is a spike in the number of cases after the introduction of the aP vaccine. A possible explanation for the observed trend is that this newer vaccine is not as effective.



Access data from the CMI-PB project

This database (like many modern projects) uses an API to return JSON format data.

We will use the R package `jsonlite`.

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE)
head(subject)
```

>Q4. How many wP (the older whole-cell vaccine) individuals and aP (newer acellular vaccine) individuals are in this database?

```{r}
sum(subject$infancy_vac == "wP")

table(subject$infancy_vac)
```

There are 60 aP infancy vaccinated subjects. There are 58 wP infancy vaccinated subjects.

>Q5. What is the number of individuals by biological sex and race?
Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$biological_sex)
table(subject$race)

table(subject$race, subject$biological_sex)
```

There are 79 female and 39 male patients in the dataset. 

The breakdown of race and biological sex is included above.

# Side-Note: Working with dates

We can use the lubridate package to ease the pain of doing math with dates.

```{r}
library(lubridate)
```
```{r}
today()
```

```{r}
today() - ymd("2000-01-01")
```

```{r}
today() - ymd("2002-05-15")
```

```{r}
today() - mdy("5-15-2002")
```

```{r}
time_length( today() - ymd("2002-05-15"),  "years")
```

>Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
wP <- data.frame(filter(subject, subject$infancy_vac == "wP"))

round(mean(wP$age))

aP <- data.frame(filter(subject, subject$infancy_vac == "aP"))

round(mean(aP$age))
```
The average age of wP individuals is 37 years old. The average age of aP individuals is 26 years old. They seem to be significantly different with a large difference in average age.

>Q8. Determine the age of all individuals at time of boost?

So what is the age of everyone on our dataset.

```{r}
subject$age <- time_length(today() - ymd(subject$year_of_birth), "years")
```

>Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(age, fill = subject$infancy_vac) +
  geom_histogram() +
  facet_wrap(vars(infancy_vac), nrow = 2)
```

I think the two different plots have barely any overlap, therefore they seem to be significantly different.

## Joining multiple tables; Get more data from CMI-PB

>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = TRUE)
head(specimen)
```

We need to **join** these two tables (subject and specimen) to make a single new "meta" table with all our metadata. We will use the `dplyr` join functions to do this.

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```

>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

Now we can read some of the other data from CMI-PB

```{r}
ab_titer <- read_json("http://cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = TRUE)
head(ab_titer)
```

One more `inner_join()` to add all our metadata in `meta` on to our `ab_data` table:

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

IgE has 6698 specimens.
IgG has 3233 specimens.
IgG1 has 7961 specimens.
IgG2 has 7961 specimens.
IgG3 has 7961 specimens.
IgG4 has 7961 specimens.

>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

There are 3 different `$dataset` values in `abdata`: 2020_dataset, 2021_dataset, and 2022_dataset.

The number of rows for the most "recent" dataset is much lower than the other two datasets.

## Examine IgG Ab titer levels

Our first exploratory plot:

```{r}
table(abdata$antigen)
```


```{r}
ggplot(abdata) +
  aes(MFI, antigen) +
  geom_boxplot()
```

Why are certain antigens and not others very variable in their detected levels here?

Can you facet or even just color by `infancy_vac`? Is there some difference?

```{r}
ggplot(abdata) +
  aes(MFI, antigen, col = infancy_vac) +
  geom_boxplot()
```

```{r}
ggplot(abdata) +
  aes(MFI, antigen, col = infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(infancy_vac), nrow = 2)
```

There are potentially some differences here but in general it is hard to tell with this whole dataset overview...

```{r}
table(abdata$dataset)
```

>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(abdata) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

There are potentially some differences here but in general it is hard to tell with this whole dataset overview.

>Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(abdata, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(abdata, antigen=="PT") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

>Q16. What do you notice about these two antigens time courses and the PT data in particular?

There does not seem to be much difference between aP or wP. However, the levels of both aP or wP are much higher than the levels of OVA.

>Q17. Do you see any clear difference in aP vs. wP responses?

I do not see a clear difference between aP and wP responses.





Let's focus in on just the 2021_dataset.

```{r}
abdata.21 <- filter(abdata, dataset == "2021_dataset")
table(abdata.21$dataset)
```

Focus on the PT antigen for IgG levels

```{r}
pt.21 <- filter(abdata.21, isotype == "IgG", antigen == "PT")
```

Plot of planned days relative to boost

```{r}
ggplot(pt.21) +
  aes(x = planned_day_relative_to_boost,
      y = MFI_normalised,
      col = infancy_vac) +
  geom_point()
```

```{r}
ggplot(pt.21) +
  aes(x = planned_day_relative_to_boost,
      y = MFI_normalised,
      col = infancy_vac,
      group = subject_id) +
  geom_point() +
  geom_line()
```

